`{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
`

# Multivariate Analysis

## Task 5

```{r}
library(foreign)
library(psych)
library(MASS)
library(car)
```

```{r}
data <- read.spss("~/studies/multivariate/data/hsb12.sav", to.data.frame = TRUE)
```

```{r}
# 1.  (10 pt.)  File hsb12.sav.  Variables   
# school= school’s id, 
# student= student’s id
# minority =1 if ethnical minority, 0 =  if not,  
# female = 1 if female, 0 if male,
# ses = social –economic status, 
# cses = centered social-economic status,
# meanses = school’s average ses
# mathach= mathematical achievements, 
# size = number of students at school,
# sector =1 for catholic school, 0 for the state school, 
# pracad = proportion of students in the academic track 
# himinty =1 if  over 40% of students are from ethnical minorities, 0 if less than 40%.

library(lme4)

# create an unconditional multilevel model with MATHACH as the dependent variable and SCHOOL as the level 2 unit.
data$SCHOOL_AND_STUDENT <- paste(data$SCHOOL, data$STUDENT %/% 2, sep = "_")
model <- lmer(MATHACH ~ 1 + (1|SCHOOL_AND_STUDENT), data = data)

summary(model)

# residual variance
residual_variance <- sigma(model)^2

# group variance
group_variance <- VarCorr(model)$SCHOOL[1]

# intraclass correlation
intraclass_correlation <- group_variance / (group_variance + residual_variance)
